---
title: Plot multiple countries on the world map
author: Yang Liu
date: '2023-05-29'
slug: plot-multiple-countries-on-the-world-map
categories:
  - Data Visualization
  - Spatial Data
tags:
  - Geodata
  - Geographic
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

This post will show it's quite easy to download and plot the administrative areas of multiple countries on the world map.
I will also showcase a bug that puzzled me for a long time and I recently figured out: strange connecting lines among countries! 

# The most straightforward way  

(You may download the world map from [Github](https://github.com/liuyanguu/Blogdown/tree/ef15e72d86d99cd538a591780822226de664e6f5/Data))

```{r}
suppressPackageStartupMessages({
library("data.table")
library("ggplot2")
library("rgdal")
library("raster")
library("rgeos")
library("here")
library("ggthemes")
})

# download data from GADM directly and row-bind the spatial polygons data frame
cnames <- c("Haiti", "Togo", "Uganda", "Ghana", "South Africa", "Angola")
download.GADM <- function(cname0) raster::getData("GADM", country = cname0, level = 1)
dt_geos <- raster::bind(lapply(cnames, download.GADM))
dt_geos <- sp::spTransform(dt_geos, CRS("+proj=robin")) # robin transformation
dt_geos_df <- broom::tidy(dt_geos, region = "GID_1")

  x_min <- min(dt_geos_df$long)*1.2
  x_max <- max(dt_geos_df$long)*1.2
  y_min <- min(dt_geos_df$lat)* 1.2
  y_max <- max(dt_geos_df$lat)* 1.2
  
  map_theme <- ggthemes::theme_map() +
    theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = c("center"))
  
  shp_world_robin <- readRDS(here::here("../Data/World.shp/sp.world.robin.rds")) # this has to be sourced locally
  ggplot() +
    geom_polygon(data = shp_world_robin, aes(x = long, y = lat, group = group), fill="lightgray", 
                 colour = "white", size=0.05) +
    geom_polygon(data = dt_geos, aes(x = long, y = lat, group = group), 
                 color = "red", size=0.05, fill="#ffc069")  +
    # if want to crop map:
    coord_fixed(xlim=c(x_min, x_max), ylim=c(y_min, y_max)) +
    map_theme +
    guides(fill = guide_legend(nrow = 1, title.position = "top"))
```

# A more complicated example  

Now we will also plot some values on the map. These values need to be merged directly to the spatial polygons _sp_ data frame, or the corresponding data frame. Here I show the second approach: transforming _sp_ data frame into a data frame first. And we can supply either the _sp_ data frame or the data frame to `ggplot2::geom_polygon`.  
  
Note that in GADM admin 1 shape files, `GID_1` is a unique identifier even after binding multiple countries, here we _intentionally_ use region names (the `NAME_1` column from GADM file) as the region identifier to show a bug. And in my case we indeed have to use the region names to merge with estimates.
  
```{r}
# `region` will become the id/group used to identify each area 
# Either broom::tidy or ggplot2::fortify can work:
# dt_geos_data <- broom::tidy(dt_geos, region = "NAME_1")
# dt_geos_data <- ggplot2::fortify(dt_geos, region = "NAME_1")

download.GADM.df <- function(cname0){
  dfsp <- raster::getData("GADM", country = cname0, level = 1) # download data from GADM
  dfsp <- sp::spTransform(dfsp, CRS("+proj=robin")) # robin transformation
  dfsf <- broom::tidy(dfsp, region = "NAME_1") # get df
  dfsf$country <- cname0 # add a country identifier 
  return(setDT(dfsf))
}

dt_geos_data <- rbindlist(lapply(cnames, download.GADM.df))

# imaging we want to plot some estimates, not only the map
dt_admin1 <- unique(setDT(dt_geos_data)[,.(country, id)])
set.seed(1234)
dt_admin1$value <- rgamma(nrow(dt_admin1), shape = 4, scale = 15) # some random values

setkey(dt_geos_data, country, id)
setkey(dt_admin1, country, id)
dt_geos_data_value <- dt_geos_data[dt_admin1]

legend_break <- c(0, 25, 50, 75,  100, 150, 200, 300, 500)
legend_label <- c("≤25", "25 to 50", "50 to 75", "75 to 100", "100 to 150","150 to 200","200 to 300",">300")
legend_color <- c("#80BD41", "#CFF4FF","#feec9f","#ffc069","#fa8c16","#d46b08","#ad4e00","#612500")
dt_geos_data_value$col <- cut(dt_geos_data_value$value, breaks = legend_break, labels = legend_label)
```

# The bug: strange connecting lines   

```{r}
ggplot() +
  geom_polygon(data = shp_world_robin, aes(x = long, y = lat, group = group), fill="lightgray", 
               colour = "white", size=0.05) +
  geom_polygon(data = dt_geos_data_value, aes(x = long, y = lat, group = group, fill= col), 
               color = "red", size=0.05)  +
  coord_fixed(xlim=c(x_min, x_max), ylim=c(y_min, y_max)) + # crop map
  scale_fill_manual("Some random values for example", values = legend_color, drop = FALSE) +# Keep all legend item
  map_theme + 
  guides(fill = guide_legend(nrow = 1, title.position = "top"))
```

  
The reason is some countries have shared admin names, which comes from `NAME_1` in the GADM file (during the `broom::tidy` or `ggplot2::fortify` step).  
Here, Haiti and Togo have these shared admin 1 names: "Centre.1" and "Centre.2" (see the table below).  
The bug is solved by setting `group` to a unique identifier in the `ggplot2::geom_polygon`:

```{r}
dt_geos_data_value[, county_group := paste(country, group, sep = "_")]
unique(dt_geos_data_value[group %in% c("Centre.1"),.(country, id, group, county_group, value)])

ggplot() +
    geom_polygon(data = shp_world_robin, aes(x = long, y = lat, group = group), fill="lightgray", 
                 colour = "white", size=0.05) +
    geom_polygon(data = dt_geos_data_value, aes(x = long, y = lat, group = county_group, fill= col), 
                 # it's important to set `group = county_group` instead of `group = group`
                 color = "red", size=0.05)  +
    coord_fixed(xlim=c(x_min, x_max), ylim=c(y_min, y_max))  +
    scale_fill_manual("Some random values for example", values = legend_color, drop = FALSE) +# Keep all legend item
    map_theme + 
    guides(fill = guide_legend(nrow = 1, title.position = "top"))
```
 
Hope this post is helpful to people who meet similar issues!   
