---
title: Working with 3D array as long-format data in R
author: Yang Liu
date: '2020-01-11'
slug: working-with-3d-array-as-long-format-data-in-r
categories:
  - Data Engineering
tags:
  - Array
output:
  blogdown::html_page:
    toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 4)
```

```{r}
check.and.install.pkgs <- function(pkgs){
  new.packages <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
  if(length(new.packages)) install.packages(new.packages, dependencies = TRUE)
  suppressPackageStartupMessages(invisible(lapply(pkgs, library, character.only = TRUE)))
}
check.and.install.pkgs(c("data.table", "reshape2", "scatterplot3d"))
```

**Happy New Year!**  

Recently I spent some time working with array in R.   
I believe it is a bad idea to work with array using `for` loop, which is both slow and error-prone. We can just melt it into a long data, do the work, and arrange back into array in the end if needed.   

## Example: a 3-D array with dimension **4x3x2**    

Notice how the values (`1:24`) fill into the three dimentions, The values `1:12` go into the 1st 4x3 matrix by column, then the rest go into the 2nd 4x3 matrix.  
```{r}
arr = array(1:24, dim=c(4,3,2), dimnames = list(group = LETTERS[1:4],  year = 2001:2003, sex = c("F", "M")))
arr
```

## Melt into long  

Melt directly, and in the long-format data, the values are sorted exactly from 1 to 24  

```{r}
arr_long <- reshape2::melt(arr)
rmarkdown::paged_table(arr_long)
```

Thus if we create an array with the same dimension, it reverts back to the original one  
```{r}
arr2 <- array(data = arr_long$value,
                     dim = c(4,3,2))
identical(unname(arr), arr2) # since I did not set dimnames for arr2, I need `unname`
```

## Recover the array  

But the problem is we usually need to work on the data and the row orders will change.  
For example, to add another year of data. And in many other cases, it is just easier to work with the long-format data.    
```{r}
arr_long_extra <- expand.grid(group = LETTERS[1:4], year = 2004, sex = c("F", "M"))
arr_long_extra$value <- 100 + 1:8
arr_long_extra
arr_long_new <- rbind(arr_long, arr_long_extra)
```

**Set the rows in the right order is the key to produce the right array**  

Since the original dimension **4x3x2** corresponds to _group_ x _year_ x _sex_, the array should be ordered by _sex_, _year_, and _group_ if we want to recover the original array. (Thus the arr_long is in the right order if we don't resort it.)

```{r}
# order by sex, year, group
arr_long_new <- arr_long_new[with(arr_long_new, order(sex, year, group)),]

# create new array with 4 groups x 4 years x 2 sex categories
# The order of the values will fill into the dimensions correctly:
arr_long_new$value
array(arr_long_new$value, dim=c(4,4,2), dimnames = list(group = LETTERS[1:4],  year = 2001:2004, sex = c("F", "M")))
```
Notice that the extra values from `arr_long_extra` have been added correctly into the array.  
  
```{r}
# Alternatively, if you use data.table, this does the same thing:
data.table::setorder(arr_long_new, sex, year, group)
```


Similarly, we can **set one dimension of the array by a specific order or using another vector.**  
Suppose I want to reorder the `year` variable as `year_order`:  
```{r}
year_order <- c("2003", "2001", "2002")
arr_long2 <- arr_long[with(arr_long, order(sex, match(year, year_order), group)),]
# notice that the new dimnames need to be assigned correctly (manually)
# as the array arranges the values correctly, there is no name automatically linked to the value
arr3 <- array(arr_long2$value, dim=c(4,3,2), dimnames = list(group = LETTERS[1:4],  year = year_order, sex = c("F", "M")))
arr3
```

It is the same as reordering the 2nd dimension of array:  
```{r}
identical(arr3,
          arr[, order(match(dimnames(arr)[[2]], year_order)),]
          )
```

To visualize the array, the long data is required as well  
```{r}
library("scatterplot3d")
shapes  <-  15:18
shapes <- shapes[as.numeric(arr_long$group)]
# haven't figured out how to round the axis annotation 
scatterplot3d(arr_long, pch = shapes, cex.symbols = 2,
              xlab = "Group", ylab = "Year", zlab = "Sex",
              highlight.3d = TRUE)
```

