---
title: How to Draw Heatmap with Colorful Dendrogram
author: Yang Liu
date: '2018-07-16'
slug: how-to-draw-heatmap-with-colorful-dendrogram
categories:
  - Data Visualization
tags:
  - R
  - Markdown
output:
  blogdown::html_page:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, 
                      fig.width=12, fig.height=12)
```

This data visualization example include:    
* Hierarchical clustering, dendrogram and heat map based on normalized odds ratios  
* The dendrogram was built separately to give color to dendrogram's branches/labels based on cluster using `dendextend`  
* Heatmap is made by `heatmap.2` from `gplots` using the built dendrogram    
* The rows are sorted by means from highest to lowest, it can be done in either the dendrogram or the `heatmap.2`    
* Use color palettes from `colorspace`  

## Data
* YRBSS (Youth Risk Behavior Surveillance System) survey data from CDC. This example plots the odds ratios of 63 behavioral questions on one question regarding disordered eating.  
* The odds ratios are median centered by column and log2 transformed

## Version 1: Color both the branches and labels
```{r}
suppressPackageStartupMessages({
library(curl)       # read file from google drive
library(gplots)     # heatmap.2
library(dendextend) # make and color dendrogram
library(colorspace) # diverge_hcl / rainbow_hcl / heat_hcl color palettes
})
id <- "1pIPphAGJcjKxkgWKrS_heLyjEPxOEYl4" # google file ID
Fasted <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))
# The ID is contained in the sharing link copied from google drive:
# https://drive.google.com/file/d/1pIPphAGJcjKxkgWKrS_heLyjEPxOEYl4/view?usp=sharing
# the dataset can be downloaded using this link as well.

names(Fasted) <- c("Questions", "2003", "2005", "2007", "2009", "2011", "2013")
# Top 5 questions and their odds ratio on Eating disorder: fasting to loose weight
head(Fasted, 5)

## Normalization, re-format into matrix ##
F_m <- as.matrix(Fasted[,2:7],dimnames=list((Fasted$Questions), names(Fasted)[2:7]))
# normalized by column median (direction: 2) and log2.
F_m2 <- (apply(F_m, 2, function(x){log2(x/median(x, na.rm = T))}))
# add the labels (Questions)
dimnames(F_m2)[1] <- list(Fasted$Questions)

## Add colors to dendrogram ##
# library(dendextend)
# library(colorspace)
# distance & hierarchical clustering
dend1 <- as.dendrogram(hclust(dist(F_m2)))
c_group <- 8 # number of clusters
dend1 <- color_branches(dend1, k = c_group, col = rainbow_hcl) # add color to the lines
dend1 <- color_labels(dend1, k = c_group, col = rainbow_hcl)   # add color to the labels

# reorder the dendrogram, must incl. `agglo.FUN = mean`
rMeans <- rowMeans(F_m2, na.rm = T)
dend1 <- reorder(dend1, rowMeans(F_m2, na.rm = T), agglo.FUN = mean)

# get the color of the leaves (labels) for `heatmap.2`
col_labels <- get_leaves_branches_col(dend1)
col_labels <- col_labels[order(order.dendrogram(dend1))]

# if plot the dendrogram alone:
# the size of the labels:
dend1 <- set(dend1, "labels_cex", 0.8)
par(mar = c(1,1,1,14))
plot_horiz.dendrogram(dend1, side = F) # use side = T to horiz mirror if needed

## plot the heatmap with the dendrogram above ##
par(cex.main=0.8)                   # adjust font size of titles
heatmap.2(F_m2, main = 'Fasting to Lose Weight',
          # reorderfun=function(d, w) reorder(d, w, agglo.FUN = mean),
                                     # order by branch mean so the deepest color is at the top
          dendrogram = "row",        # no dendrogram for columns
          Rowv = dend1,              # * use self-made dendrogram
          Colv = "NA",               # make sure the columns follow data's order
          col = diverge_hcl,         # color pattern of the heatmap
          
          trace="none",              # hide trace
          density.info="none",       # hide histogram
          
          margins = c(5,18),         # margin on top(bottom) and left(right) side.
          cexRow=1, cexCol = 0.8,      # size of row / column labels
          xlab = "Year",
          srtCol=0, adjCol = c(0.5,1), # adjust the direction of row label to be horizontal
          # margin for the color key
          # ("bottom.margin", "left.margin", "top.margin", "left.margin" )
          key.par=list(mar=c(5,1,3,1)),
          RowSideColors = col_labels, # to add nice colored strips        
          colRow = col_labels         # add color to label
          )
```

## Version 2: color only the labels.
* There is no need to use the predesigned `dend1` in `heatmap.2`  
* But then need to reorder the trees by adding `reorderfun`  
* Still need to create the `col_labels`  
```{r}
heatmap.2(F_m2, main = 'Fasting to Lose Weight',
          reorderfun=function(d, w) reorder(d, w, agglo.FUN = mean),
                                     # order by branch mean so the deepest color is at the top
          dendrogram = "row",        # no dendrogram for columns
          # Rowv = dend1,              # * use self-made dendrogram
          Colv = "NA",               # make sure the columns follow data's order
          col = diverge_hcl,         # color pattern of the heatmap
          
          trace="none",              # hide trace
          # density.info="none",       # hide histogram
          
          margins = c(5,18),         # margin on top(bottom) and left(right) side.
          cexRow=1, cexCol = 0.8,      # size of row / column labels
          xlab = "Year",
          srtCol=0, adjCol = c(0.5,1), # adjust the direction of row label to be horizontal
          # margin for the color key
          # ("bottom.margin", "left.margin", "top.margin", "left.margin" )
          key.par=list(mar=c(5,1,3,1)),
          # RowSideColors = col_labels, # to add nice colored strips        
          colRow = col_labels         # add color to label
          )
```

## Version 3: If there is no color, and we do not reorder the branches
* Then there is no need to create an extra dendrogram using `dendextend`
```{r}
par(cex.main=0.8)                   # adjust font size of titles
heatmap.2(F_m2, main = 'Fasting to Lose Weight',
          # reorderfun=function(d, w) reorder(d, w, agglo.FUN = mean),
                                     # order by branch mean so the deepest color is at the top
          dendrogram = "row",        # no dendrogram for columns
          Colv = "NA",               # make sure the columns follow data's order
          col = diverge_hcl,         # color pattern of the heatmap
          trace="none",              # hide trace
          density.info="none",       # hide histogram
          margins = c(5,18),         # margin on top(bottom) and left(right) side.
          cexRow=1, cexCol = 0.8,      # size of row / column labels
          xlab = "Year",
          srtCol=0, adjCol = c(0.5,1), # adjust the direction of row label to be horizontal
          # margin for the color key
          # ("bottom.margin", "left.margin", "top.margin", "left.margin" )
          key.par=list(mar=c(5,1,3,1))
          )
```
