---
title: "Study shrinkage and DART in xgboost modeling using a simple dataset"
author: Yang Liu
date: "2018-11-15"
slug: xgboost-dart-example
categories:
  - Machine Learning
tags:
  - XGBoost
  - DART
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r loadlibs, include = F}
suppressPackageStartupMessages({
library(xgboost)
library(DiagrammeR)
})
options(digits = 4)
```

It is always a good idea to study the packaged algorithm with a simple example. Inspired by [my colleague Kodi's excellent work showing how `xgboost` handles missing values](http://arfer.net/w/xgboost-sparsity), I tried a simple 5x2 dataset to show how shrinkage and DART influence the growth of trees in the model.

# Data
```{r}
set.seed(123)
n0 <- 5
X <-  data.frame(x1 = runif(n0), x2 = runif(n0))
Y <-  c(1, 5, 20, 50, 100)
cbind(X, Y)
```

# Shrinkage
1. Step size shrinkage was the major tool designed to prevents overfitting (over-specialization).   
* [The R document](https://xgboost.readthedocs.io/en/latest/parameter.html) says that the learning rate `eta` has range [0, 1] but `xgboost` takes any value of $eta\ge0$. Here I select eta = 2, then the model can perfectly predict in two steps, the train rmse from iter 2 was 0, only two trees were used.   
Of course, it is a bad idea to use a very large _eta_ in real applications as the tree will not be helpful and the predicted value will be very wrong.   
* The `max_depth` is the maximum depth of a tree, I set 10 but it won't be reached    
* By default, there is no other regularization
* By setting `base_score = 0` in `xgboost` we can add up the values in the leaves of two trees to get every number in Y: `r Y`
```{r}
# non-zero skip_drop has higher priority than rate_drop or one_drop
param_gbtree <- list(objective = 'reg:linear', nrounds = 3, 
                   eta = 2,
                   max_depth = 10,
                   min_child_weight = 0,
                   booster = 'gbtree'
)

simple.xgb.output <- function(param,...){
  set.seed(1234)
  m = xgboost(data = as.matrix(X), label = Y, params = param,
              nrounds = param$nround, 
              base_score = 0)
  cat('Evaluation log showing testing error:\n')
  print(m$evaluation_log)
  pred <- predict(m, as.matrix(X), ntreelimit = param$nrounds)
  cat('Predicted values of Y: \n')
  print(pred)
  pred2 <- predict(m, as.matrix(X), predcontrib = TRUE)
  cat("SHAP value for X: \n")
  print(pred2)
  p <- xgb.plot.tree(model = m)
  p
}
simple.xgb.output(param_gbtree)
```
   
* If `eta` = 1  
Then no perfect prediction could be made and the trees grow in a more conservative manner  
```{r}
# non-zero skip_drop has higher priority than rate_drop or one_drop
param_gbtree <- list(objective = 'reg:linear', nrounds = 3, 
                   eta = 1,
                   max_depth = 10,
                   min_child_weight = 0,
                   booster = 'gbtree'
)
simple.xgb.output(param_gbtree)
```

# DART: Dropout - MART
DART ([paper on JMLR](http://proceedings.mlr.press/v38/korlakaivinayak15.pdf)) adopted dropout method from neural networks to boosted regression rees (i.e.,MART: Multiple Additive Regression Trees). DART aims to further prevent over-specialization. It requires select `booster = 'dart'` in `xgboost` and tune several hyper-parameters ([Ref: R documents](https://xgboost.readthedocs.io/en/latest/parameter.html#additional-parameters-for-dart-booster-booster-dart)).  
  
## `skip_drop`  
* The `skip_drop`(default = 0, range [0, 1]) is the probability of skipping dropout. It has a higher priority than other DART parameters. If `skip_drop` = 1, the dropout procedure would be skipped and `dart` is the same as `gbtree`. The setting below gives the same result as the `gbtree` above (results omitted):
```{r}
param_gbtree <- list(objective = 'reg:linear', nrounds = 3, 
                   eta = 2,
                   max_depth = 10,
                   booster = 'dart',
                   skip_drop = 1,  # = 1 means always skip, = gbtree
                   rate_drop = 1,  # doesn't matter since drop is always skipped
                   one_drop = 1
)
```
  
## `rate_drop`  
* If `skip_drop`$\ne0$, `rate_drop (default = 0, range [0, 1])` will drop a fraction of the trees before the model update in every iteration. 
* The DART paper [JMLR](http://proceedings.mlr.press/v38/korlakaivinayak15.pdf) said the dropout makes DART between gbtree and random forest: "If no tree is dropped, DART is the same as MART (`gbtree`); if all the trees are dropped, DART is no different than random forest."
* If `rate_drop` = 1 then all the trees are dropped, a random forest of trees is built. In our case of a very simple dataset, the 'random forest' just repeats the same tree `nrounds` times:   
```{r}
param_dart1 <- list(objective = 'reg:linear', nrounds = 3, 
                   eta = 2,
                   max_depth = 10,
                   booster = 'dart',
                   skip_drop = 0,  
                   rate_drop = 1,  # doesn't matter since drop is always skipped
                   one_drop = 1
)
simple.xgb.output(param_dart1)
```
  
## `one_drop`  
* If `one_drop` = 1 then at least one tree is always dropped. If I let `rate_drop`=0, but `one_drop` = 1, the dropping was still working, and the trees were built in a more conservative manner. Since the first tree will be dropped, the second tree is the same as the first one  
```{r, fig.width = 8, fig.height = 10}
param_dart2 <- list(objective = 'reg:linear', nrounds = 5, 
                   eta = 2,
                   max_depth = 10,
                   booster = 'dart',
                   skip_drop = 0,  
                   rate_drop = 0,  # doesn't matter since drop is always skipped
                   one_drop = 1
)
simple.xgb.output(param_dart2)
```

* Similar conservative effect if I set `skip_drop` to be non-zero:  
```{r, fig.width = 8, fig.height = 10}
param_dart3 <- list(objective = 'reg:linear', nrounds = 5, 
                   eta = 2,
                   max_depth = 10,
                   booster = 'dart',
                   skip_drop = 0,  
                   rate_drop = 0.5,  # doesn't matter since drop is always skipped
                   one_drop = 0
)
simple.xgb.output(param_dart3)
```

Letting `one_drop` = 1 also gives result more conservative, and smaller train-rmse if use same rounds. 