---
title: 'SHAP Visualization in R (first post)'
author: Yang Liu
date: '2018-10-14'
slug: shap-visualization-for-xgboost
categories:
  - Machine Learning
  - Data Visualization
tags:
  - XGBoost
  - SHAP
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

Update 19/07/21:

Since my R Package **SHAPforxgboost** has been released on CRAN, I updated this post using the new functions and illustrate how to use these functions using two datasets. For more information, please refer to: [SHAP visualization for XGBoost in R](https://liuyanguu.github.io/post/2019/07/18/visualization-of-shap-for-xgboost/)

# Example 1

This is the example I used in the package **SHAPforxgboost**

```{r}
# Example use iris
suppressPackageStartupMessages({
library(SHAPforxgboost)
library(xgboost)
library(data.table)
library(ggplot2)
})

X1 = as.matrix(iris[,-5])
mod1 = xgboost::xgboost(
  data = X1, label = iris$Species, gamma = 0, eta = 1, 
  lambda = 0,nrounds = 1, verbose = F)

# shap.values(model, X_dataset) returns the SHAP
# data matrix and ranked features by mean|SHAP|
shap_values <- shap.values(xgb_model = mod1, X_train = X1)
shap_values$mean_shap_score
shap_values_iris <- shap_values$shap_score

# shap.prep() returns the long-format SHAP data from either model or
shap_long_iris <- shap.prep(xgb_model = mod1, X_train = X1)
# is the same as: using given shap_contrib
shap_long_iris <- shap.prep(shap_contrib = shap_values_iris, X_train = X1)
```

## SHAP summary plot

```{r}
shap.plot.summary(shap_long_iris)
# option of dilute is offered to make plot faster if there are over thousands of observations
# please see documentation for details. 
shap.plot.summary(shap_long_iris, x_bound  = 1.5, dilute = 10)
```

## Alternative ways:

```{r, eval = F}
# option 1: from the xgboost model
shap.plot.summary.wrap1(mod1, X1, top_n = 3)

# option 2: supply a self-made SHAP values dataset (e.g. sometimes as output from cross-validation)
shap.plot.summary.wrap2(shap_score = shap_values$shap_score, X1, top_n = 3)
```

## SHAP dependence plot

```{r}
shap.plot.dependence(data_long = shap_long_iris, x="Petal.Length",
                           y = "Petal.Width", color_feature = "Petal.Width")
```

The without color version, just plot SHAP value against feature value:

```{r}
shap.plot.dependence(data_long = shap_long_iris, "Petal.Length")
```

## SHAP interaction effect plot

```{r}
# To get the interaction SHAP dataset for plotting:
# fit the xgboost model
mod1 = xgboost::xgboost(
  data = as.matrix(iris[,-5]), label = iris$Species,
  gamma = 0, eta = 1, lambda = 0,nrounds = 1, verbose = FALSE)
# Use either:
data_int <- shap.prep.interaction(xgb_mod = mod1,
                                  X_train = as.matrix(iris[,-5]))
# or:
shap_int <- predict(mod1, as.matrix(iris[,-5]),
                    predinteraction = TRUE)

# **SHAP interaction effect plot **
shap.plot.dependence(data_long = shap_long_iris,
                           data_int = shap_int_iris,
                           x="Petal.Length",
                           y = "Petal.Width",
                           color_feature = "Petal.Width")


```

## SHAP force plot

```{r}

# **SHAP force plot**
plot_data <- shap.prep.stack.data(shap_contrib = shap_values_iris,
                                  n_groups = 4)
shap.plot.force_plot(plot_data)
shap.plot.force_plot(plot_data, zoom_in_group = 2)

# plot all the clusters:
shap.plot.force_plot_bygroup(plot_data)

```

# Example 2

This example is based on the slum data I used in the earilier post.

```{r libs, include = FALSE, echo = FALSE}
# Data Preparation using data.table -----------------------------
# https://drive.google.com/open?id=1pHE8ktykKfBN2_yo8FNxYX_BFKp1S0SN
id <- "1pHE8ktykKfBN2_yo8FNxYX_BFKp1S0SN" # google file ID
mydata <- fread(sprintf("https://docs.google.com/uc?id=%s&export=download", id),
                stringsAsFactors = TRUE)

# clean variables names, remove "/".
names(mydata) <- gsub("/", "", names(mydata)) 

# create variable: Share_Temporary
# The dependent variable to be modeled: Share of temporary structure in slums
mydata[, Share_Temporary := CC7_Structures_Temporary / (CC7_Structures_Temporary + CC6_Structures_Permanent)]

# remove variables with over 20% NA, var_keeps those is.na < 20%
var_keep <- sapply(mydata, function(x) sum(is.na(x))) < dim(mydata)[1]*0.2
var_keep[length(var_keep)] <- TRUE # keep "Share_Temporary"
mydata[, names(mydata)[!var_keep]:= NULL]
mydata <- na.omit(mydata) # remove rest N.A.
# fix some format issue
Factor_List <- c("GG4_Toilets_Pay_Amount", "CC11_Population_Estimate", "CC10_Household_Size",
                 "CC12_Total_Population")
mydata[, (Factor_List) := lapply(.SD, as.numeric), .SDcols = Factor_List]

# remove useless factor
mydata[, c("Obs. Count", "City", "Country",
                       names(mydata)[grepl("Structure", names(mydata))],
                       names(mydata)[grepl("Ownership", names(mydata))]):=NULL]
# remove what are highly correlated to dependent variable

# divide training and test dataset
set.seed(123)
train_idx <- sample(dim(mydata)[1], dim(mydata)[1]* 0.6)

# The model.matrix() function is used in many regression packages for building 
# an "X" matrix from data.
# need matrix for glmnet
X2 <- model.matrix(Share_Temporary~., data = mydata)
Y2 <- as.matrix(mydata[,"Share_Temporary"])
X_train <- X2[train_idx,]
X_test <- X2[-train_idx,]
Y_train <- Y2[train_idx]
Y_test <- Y2[-train_idx]
# merge back to df again, as df is required for regsubsets (best subset)
data_train <- data.frame(X2, Y2)[train_idx,] 
data_test <- data.frame(X2, Y2)[-train_idx,] 
```

## Summary plot

-   Using `geom_sina` from `ggforce` to make the sina plot\
-   We can see clearly for the most influential variable on the top: Monthly water cost. A Higher cost is associated with the declined share of temporary housing. But a very low cost has a strong impact on the increased share of temporary housing\
-   The effects of binary variables are highly distinctive. The second variable shows that Resettled housing is highly unlikely to be temporary, so does being close to wells as water sources.

**Load the xgboost model**

```{r}
best_rmse_index <- 56
best_rmse <- 0.2102
best_seednumber <- 3660
best_param <- list(objective = "reg:linear",  # For regression
                eval_metric = "rmse",      # rmse is used for regression
                max_depth = 9,
                eta = 0.09822,   # Learning rate, default: 0.3
                subsample = 0.64,
                colsample_bytree = 0.6853, 
                min_child_weight = 6, # These two are important
                max_delta_step = 8)
# The best index (min_rmse_index) is the best "nround" in the model
nround <- best_rmse_index
set.seed(best_seednumber)
mod1 <- xgboost::xgboost(data = X_train, label = Y_train, params = best_param, nround = nround, verbose = F)

```

```{r}
shap.plot.summary.wrap1(mod1, X_train, top_n = 10)
```

## Dependence plot for each feature

-   Here we choose to show top 6 features ranked by mean\|SHAP\|\

```{r, fig.width=8, fig.height=6}
data_long <-  shap.prep(mod1, X_train = X_train)
shap_values <- shap.values(mod1, X_train)
features_ranked <- names(shap_values$mean_shap_score)[1:4]
fig_list <- lapply(features_ranked, shap.plot.dependence, data_long = data_long)
gridExtra::grid.arrange(grobs = fig_list, ncol = 2)
```

-   If Use the built-in `xgb.shap.plot` function\

```{r, echo = T , fig.width=8, fig.height=6}
xgboost::xgb.plot.shap(data = X_train, model = mod1, top_n = 4, n_col = 2)

```

## Force plot

Since there are so many features in this dataset, we pick only top 6 and merge the rest.

-   Use `geom_col` to show features each contributing to push the model output from the base value (the average model output) to the model output.\
-   Have tried geom_area but dones't work very well due to gaps in the plot caused by fluctuation of positive and negative values.\
-   apply the order of clustering to group observations under similar influences closer.

```{r}
force_plot_data <- shap.prep.stack.data(shap_contrib = shap_values$shap_score, top_n  = 5, n_groups = 4)
shap.plot.force_plot(force_plot_data)

```

## Stack plot by clustering groups

```{r}
shap.plot.force_plot_bygroup(force_plot_data)

```
